Analysing the effects of COVID-19 on Crime in the United States

### 2022-04-30 outputs/paper/paper.Rmd
Added text for Seattle and Philly with details on data sources and overview of exploratory analysis.

```{r graph5, fig.cap = "Daily Confirmed Cases of COVID-19 in L.A",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 4, fig.width = 8,}
# 2020 only
knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
copy  = la|>mutate(date_occ=as.Date(date_occ))

@@

```
Los Angeles, located in California, has historically been a hub of criminal activities mainly centered around narcotics due to its proximity to Tijuana and in general the US-Mexico border. Crime in Los Angeles is also prevalent due to the presence of one of the largest populations of homeless people in the United States mostly within the Skid-Row locality. Los Angeles was at the peak of criminal activity between 1970 and 1990, and more recently in 2015, it was revealed that the Los Angeles Police were under-reporting criminal activity for roughly 8 years, artificially lowering the crime rates and making it an intriguing city to examine from a data perspective [@bird2018impact]. To examine the crime rates within the city we source data directly from the  [Los Angeles Police Department Socrata API](https://dev.socrata.com/foundry/data.lacity.org/2nrs-mtv8) and look at the top five crimes which are mainly variants of burglary and theft crimes, alongside vehicle theft and assault. Finally, as shown in Figure \@ref(fig:graph5), the pandemic had a significant impact in Los Angeles with some of the first patients of COVID-19, including "patient zero", being from Los Angeles. California overall also faced a severe health crisis with cases up to ten times more than were anticipated.

@@

daily
```
<!-- \newpage -->

@@

Nulla consectetur erat arcu, eu porta urna sagittis nec. In sed vestibulum turpis. Nullam vel libero eu sapien eleifend condimentum. Donec nunc mi, ornare quis viverra ut, volutpat vitae orci. Quisque sit amet porttitor mauris, faucibus maximus arcu. Duis et lorem eu elit consectetur posuere. Aliquam blandit nisl nisl. Proin velit arcu, gravida vel tellus.
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec egestas justo in nulla bibendum ultrices. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec porttitor interdum purus, vel commodo lorem lacinia at. Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin. Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar. Proin sapien elit, tempor non consectetur ut, mollis quis nunc. Maecenas tincidunt velit dolor, eu porta lacus ornare et. Ut in cursus felis. Etiam porta, ipsum sit amet pulvinar consectetur, nulla metus vehicula quam, ac blandit lacus ex et ante. Aliquam lacinia sagittis lobortis. Integer eros felis, gravida nec augue ut, blandit ultricies ipsum. Morbi aliquet at magna rhoncus vehicula. Suspendisse aliquam risus nunc, vitae eleifend lacus venenatis ut. Sed nisi eros, malesuada nec purus ut, tristique congue dui.Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec egestas justo in nulla bibendum ultrices. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec porttitor interdum purus, vel commodo lorem lacinia at. Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin. Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar. Proin sapien elit, tempor non consectetur ut, mollis quis nunc. Maecenas tincidunt velit dolor, eu porta lacus ornare et. Ut in cursus felis. Etiam porta, ipsum sit amet pulvinar consectetur, nulla metus vehicula quam, ac blandit lacus ex et ante. Aliquam lacinia sagittis lobortis. Integer eros felis, gravida nec augue ut, blandit ultricies ipsum. Morbi aliquet at magna rhoncus vehicula. Suspendisse aliquam risus nunc, vitae eleifend lacus venenatis ut. Sed nisi eros, malesuada nec purus ut, tristique congue dui.

```{r graph10, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}

@@

# try first log difference
ts_diff_PH <- na.omit(diff(ts_PH))
plot.xts(ts_diff_PH,
         main = "Daily confirmed cases of COVID-19 in Philadelphia")
```

@@

>>>>

```{r graph5, fig.cap = "Daily Confirmed Cases of COVID-19 in L.A",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 3, fig.width = 8,}
# 2020 only
knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
copy  = la|>mutate(date_occ=as.Date(date_occ))

@@

```
\newpage

Los Angeles, located in California, has historically been a hub of criminal activities mainly centered around narcotics due to its proximity to Tijuana and in general the US-Mexico border. Crime in Los Angeles is also prevalent due to the presence of one of the largest populations of homeless people in the United States mostly within the Skid-Row locality. Los Angeles was at the peak of criminal activity between 1970 and 1990, and more recently in 2015, it was revealed that the Los Angeles Police were under-reporting criminal activity for roughly 8 years, artificially lowering the crime rates and making it an intriguing city to examine from a data perspective [@bird2018impact].

To examine the crime rates within the city we source data directly from the  [Los Angeles Police Department Socrata API](https://dev.socrata.com/foundry/data.lacity.org/2nrs-mtv8) and look at the top five crimes which are mainly variants of burglary and theft crimes, alongside vehicle theft and assault. Finally, as shown in Figure \@ref(fig:graph5), we note that the pandemic had a significant impact in Los Angeles with some of the first patients of COVID-19, including "patient zero", being from Los Angeles. California overall also faced a severe health crisis with cases up to ten times more than were anticipated.

@@

daily
```
\newpage

@@

One of the fastest-growing cities in the United States, Seattle, was specifically chosen for this study due to its high rates of property crime while having relatively constant overall crime rates across the decades.  As shown in Figure \@ref(fig:graph8), the city was relatively less impacted by COVID-19, making it an interesting candidate in contrast to other cities when it comes to observing criminal behavior. Seattle is also the location of the earliest clinical trial being conducted for the Moderna COVID-19 vaccine which occurred around March 20, 2020 [@tuesday_2020], and more recently has seen a sudden increase in criminal activity alongside increased homelessness [@westneat_2022].

 The crime data for Seattle was sourced from the public [Seattle Police Department Records](https://data.seattle.gov/Public-Safety/SPD-Crime-Data-2008-Present/tazs-3rd5), with our focus mainly being on the top crimes comprising assault, burglary, vandalism, larceny and theft from a vehicle. Given the relatively low rate of COVID-19 in Seattle, trends across most crimes seemed to remain constant before and after the onset of the pandemic as seen in Figure \@ref(fig:graph9). It was also observed that crimes such as burglary increased while theft from motor vehicles showed a decline as compared to previous trends.  These relatively consistent trends during the pandemic are also visible in Figure \@ref(fig:graph10), with an increase in larceny, burglary, and vandalism crimes noted in June-July 2020 due to the Goerge Floyd protests.

```{r graph10, fig.cap = "Frequency of each crime in Seattle during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}

@@

# try first log difference
ts_diff_PH <- na.omit(diff(ts_PH))
plot.xts(ts_diff_PH,
         main = "Daily confirmed cases of COVID-19 in Philadelphia")
```
The last city in our study is Philadelphia, which is one of the most violent cities in America along with the highest poverty rates. The city is the sixth most populous city in the United States with a rich cultural diversity making it ideal for our study. The city was significantly affected by the pandemic as noted by its poor economic performance and decline in the supply of workers [@haider_eichel].  It should be noted though that when compared to the other cities in our study, Philadelphia has the lowest number of daily cases on average as shown in Figure Figure \@ref(fig:graph11).

To examine the impact on crime due to the pandemic we sourced data from the [Philadelphia Police Department](https://www.phillypolice.com/crime-maps-stats/), with an emphasis on the top crimes which were basically assaults, other offenses, thefts from vehicles, general thefts and vandalism. Overall it was found, as demonstrated in Figure \@ref(fig:graph12) that petty offenses increased during the pandemic while most other crimes remained relatively consistent with patterns emerging in prior years.  An immediate downtrend was also observed in crimes such as assaults and thefts right after the onset of the pandemic.

@@

Further examining the crime trends during years spanning the pandemic, as shown in Figure \@ref(fig:graph13), we observe that there is a general downtrend in petty offenses while most other crimes remain consistent with some seasonal trends in assaults and thefts.  Finally, as noted previously for other cities, an uptick in criminal activity is observed during the black lives matter protests of June 2020.
